# What a microsample looks like {#data}

Before any spatial analysis, we need a community table we can trust. This
chapter builds one from raw mapping output and shows why micro-scale data needs
a filtering step that ordinary metagenomics can skip.

## The raw tables

Reads from each microsample were mapped against the catalogue of 223 genomes.
That produces two tables of exactly the same shape.

```{r peek_counts, warning = FALSE, message = FALSE}
# how many reads from each microsample mapped to each genome
genome_read_counts[1:5, 1:4]

# how many bases of each genome those reads covered
genome_covered_bases[1:5, 1:4]

dim(genome_read_counts)   # genomes x (1 + microsamples)
```

And a table describing the microsamples themselves:

```{r peek_metadata, warning = FALSE, message = FALSE}
sample_metadata %>%
  select(microsample, cryosection, region, Xcoord, Ycoord, distance_host,
         after_filtering.total_bases) %>%
  head()
```

`Xcoord` and `Ycoord` are the position of the microsample on the cryosection in
micrometres. `distance_host` is how far it sits from the epithelium — the gut
wall — which is the gradient we will eventually be testing against.

::: {.rmdnote}
**Scale check.** A microsample covers a few hundred micrometres. The whole
cryosection is a couple of millimetres across. The 168 microsamples of one
section together weigh less than a grain of salt.
:::

## Sequencing depth is data here, not a nuisance

The introduction described the unnormalised pooling protocol. You can verify it
in the metadata: every microsample was cut to the same area and given the same
number of PCR cycles.

```{r design_check, warning = FALSE, message = FALSE}
sample_metadata %>%
  count(cryosection, size, cycles, name = "microsamples")
```

Identical input area, identical amplification, no equimolar normalisation. So
how much do the resulting library sizes differ?

```{r depth_range, warning = FALSE, message = FALSE}
sample_metadata %>%
  group_by(cryosection) %>%
  summarise(
    total_Gb   = round(sum(after_filtering.total_bases) / 1e9, 1),
    min_Mb     = round(min(after_filtering.total_bases) / 1e6, 1),
    median_Mb  = round(median(after_filtering.total_bases) / 1e6, 1),
    max_Mb     = round(max(after_filtering.total_bases) / 1e6, 1),
    fold_range = round(max(after_filtering.total_bases) /
                       min(after_filtering.total_bases)),
    .groups = "drop"
  )
```

A **400- to 1,400-fold** spread, between pieces of tissue of identical size
processed identically. Under equimolar pooling this range would have been
flattened to near-uniform by design. Here it survives, and the most plausible
thing it reflects is how much microbial biomass each microsample contained.

```{r depth_dist, warning = FALSE, message = FALSE, fig.height = 3.5, fig.width = 7}
ggplot(sample_metadata,
       aes(x = after_filtering.total_bases, fill = region)) +
  geom_histogram(bins = 40, alpha = 0.85, position = "identity") +
  scale_x_log10(labels = scales::label_number(scale = 1e-6, suffix = " Mb")) +
  scale_fill_manual(values = region_colors) +
  labs(x = "Sequenced bases per microsample (log scale)",
       y = "Microsamples", fill = "Region")
```

Keep this in mind for the rest of the practical. When sequencing depth turns up
as a strong predictor in Chapter \@ref(space), the reflex from ordinary
metagenomics — *that is a technical artefact, control for it* — is the wrong
reflex. Under this protocol it is a biological quantity in disguise.

It is not a clean one, though. Depth also determines how many genomes you can
possibly detect, so richness and depth are confounded by construction. We will
see exactly how strongly in a moment.

## Why micro-scale data needs a coverage filter

Each microsample contains very little DNA, so it is sequenced shallowly. Look
at how many genomes appear to be present in a typical microsample if we simply
count non-zero entries:

```{r naive_richness, warning = FALSE, message = FALSE}
naive_present <- colSums(genome_read_counts[, -1] > 0)

summary(naive_present)
```

Almost every genome in the catalogue picks up *some* reads in *some*
microsample. That cannot all be real. Short read mapping is a heuristic: a read
from a genome that is genuinely present will also align, imperfectly, to its
relatives. Those stray reads land scattered across the wrong genome.

That scattering is what lets us tell the two cases apart. A genome that is
really there gets covered **evenly along its length**; a genome that only
catches stray reads gets a few isolated patches. So instead of asking *how many
reads*, we ask *how much of the genome was covered*.

```{r coverage, warning = FALSE, message = FALSE}
genome_coverage <- breadth_of_coverage(genome_covered_bases, genome_metadata)

genome_coverage[1:5, 1:4]
```

```{r coverage_hist, warning = FALSE, message = FALSE, fig.height = 4, fig.width = 7}
coverage_long <- genome_coverage %>%
  pivot_longer(-genome, names_to = "microsample", values_to = "coverage") %>%
  filter(coverage > 0)

ggplot(coverage_long, aes(x = coverage)) +
  geom_histogram(bins = 60, fill = "#4a6ab7", colour = NA) +
  geom_vline(xintercept = 0.3, linetype = "dashed", colour = "#E76F51",
             linewidth = 0.8) +
  annotate("text", x = 0.33, y = Inf, label = "  0.3 threshold",
           hjust = 0, vjust = 2, colour = "#E76F51", size = 3.5) +
  scale_y_log10() +
  labs(x = "Fraction of genome length covered",
       y = "Genome × microsample pairs (log scale)",
       title = "Breadth of coverage is strongly bimodal")
```

The distribution has two humps: a large one near zero (genomes catching a
handful of stray reads) and a smaller one at high coverage (genomes genuinely
present). Almost nothing sits in between, which is what makes a threshold
defensible. We use **0.3**, as in the published analysis.

## Filtering and normalising

Two operations turn the raw table into a community table.

**Coverage filtering** zeroes out any genome–microsample pair below the
threshold:

```{r filtering, warning = FALSE, message = FALSE}
read_counts_filt <- filter_by_coverage(genome_read_counts, genome_coverage,
                                       min_coverage = 0.3)

tibble(
  step = c("before filtering", "after filtering"),
  nonzero_entries = c(sum(genome_read_counts[, -1] > 0),
                      sum(read_counts_filt[, -1] > 0)),
  reads_kept_pct = c(100, round(100 * sum(read_counts_filt[, -1]) /
                                  sum(genome_read_counts[, -1]), 1))
)
```

Note what happened: most *entries* are discarded, but almost all *reads* are
kept. The filter removes a huge number of near-empty cells that were carrying
very little signal.

**Length normalisation** converts read counts into genome counts. A 6 Mb genome
captures twice as many reads as a 3 Mb genome at equal abundance, so we divide
by genome length expressed in reads:

```{r genome_counts, warning = FALSE, message = FALSE}
genome_counts <- to_genome_counts(read_counts_filt, genome_metadata,
                                  read_length = 150)

genome_counts[1:5, 1:4]
```

::: {.rmdnote}
**Why not rarefy?** Rarefying subsamples every microsample down to a common
depth. That would throw away exactly the quantity this protocol went to trouble
to preserve — the depth differences that track biomass — and it would still
leave the *within*-microsample counts compositional. Chapter \@ref(space)
instead splits the problem in two: a log-ratio transformation for what the
community is made of, and depth kept as its own variable for how much of it
there was.
:::

## Diversity per microsample

With a community table we can ask how diverse each microsample is. We use Hill
numbers, which put richness and evenness on the same scale (effective number of
genomes):

```{r diversity, warning = FALSE, message = FALSE}
diversity <- alpha_diversity(genome_counts, genome_tree)

sample_metadata <- sample_metadata %>%
  left_join(diversity, by = "microsample")

sample_metadata %>%
  group_by(region) %>%
  summarise(
    microsamples  = n(),
    richness      = median(richness),
    neutral       = round(median(neutral), 1),
    phylogenetic  = round(median(phylogenetic), 1),
    .groups = "drop"
  )
```

| Metric | q | Reads as |
|---|---|---|
| `richness` | 0 | how many genomes are present |
| `neutral` | 1 | how many genomes are *effectively* present, weighting by abundance |
| `phylogenetic` | 1 | same, but close relatives count as partly redundant |

Richness far above neutral diversity means a microsample is dominated by a few
genomes with a long rare tail — which is exactly what these communities look
like.

And here is the confounding promised above:

```{r depth_richness, warning = FALSE, message = FALSE, fig.height = 3.5, fig.width = 7}
sample_metadata %>%
  filter(richness > 0) %>%
  ggplot(aes(x = after_filtering.total_bases, y = richness, colour = region)) +
  geom_point(size = 1.3, alpha = 0.75) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.7) +
  scale_x_log10(labels = scales::label_number(scale = 1e-6, suffix = " Mb")) +
  scale_colour_manual(values = region_colors) +
  labs(x = "Sequenced bases (log scale)", y = "Richness", colour = "Region")
```

```{r depth_richness_cor, warning = FALSE, message = FALSE}
sample_metadata %>%
  filter(richness > 0) %>%
  group_by(region) %>%
  summarise(spearman_rho = round(cor(richness,
                                     log(after_filtering.total_bases),
                                     method = "spearman"), 2),
            .groups = "drop")
```

ρ above 0.9. Two mechanisms produce that, and this dataset cannot separate them:
a microsample with more biomass genuinely holds more genomes, **and** a
microsample sequenced more deeply detects more of whatever it holds. Both are
real; neither can be switched off. Any claim resting on richness alone in this
kind of data is on thin ice.

```{r diversity_plot, warning = FALSE, message = FALSE, fig.height = 4, fig.width = 7}
sample_metadata %>%
  pivot_longer(c(richness, neutral, phylogenetic),
               names_to = "metric", values_to = "diversity") %>%
  mutate(metric = factor(metric,
                         levels = c("richness", "neutral", "phylogenetic"))) %>%
  ggplot(aes(x = region, y = diversity, fill = region)) +
  geom_boxplot(outlier.size = 0.6, width = 0.6, alpha = 0.85) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = region_colors, guide = "none") +
  labs(x = NULL, y = "Diversity (effective number of genomes)")
```

## Who is actually there

```{r composition, warning = FALSE, message = FALSE, fig.height = 5, fig.width = 9}
composition <- genome_counts %>%
  pivot_longer(-genome, names_to = "microsample", values_to = "count") %>%
  group_by(microsample) %>%
  mutate(rel = count / sum(count)) %>%
  ungroup() %>%
  left_join(genome_metadata %>% select(genome, phylum, order),
            by = "genome") %>%
  left_join(sample_metadata %>% select(microsample, region, distance_host),
            by = "microsample") %>%
  filter(rel > 0)

composition %>%
  # order microsamples by distance to the gut wall, the gradient of interest
  mutate(microsample = fct_reorder(microsample, distance_host)) %>%
  ggplot(aes(x = microsample, y = rel, fill = order)) +
  geom_col(width = 1) +
  facet_grid(~ region, scales = "free_x", space = "free_x") +
  scale_fill_manual(values = order_colors) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(x = "Microsample (ordered by distance to gut wall →)",
       y = "Relative abundance", fill = "Order") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        legend.key.size = unit(0.35, "cm"), legend.text = element_text(size = 7))
```

Two things are already visible, and they set up the rest of the practical:

1. **Caecum and colon are different communities.** No surprise — they are
   different organs.
2. **Within each section, neighbouring bars resemble each other.** Microsamples
   are sorted here by distance to the gut wall, and the composition does not
   look randomly shuffled along that axis.

Point 2 is a hint, not evidence. A barplot sorted by a variable will always
look somewhat ordered. Chapter \@ref(space) tests it properly.
