# Is the community spatially structured? {#space}

Chapter \@ref(data) produced a community table. This chapter asks the question
that makes the data worth collecting: **do nearby microsamples host more similar
communities than distant ones?**

We will attack it with the standard microbial ecology toolkit — ordination,
PERMANOVA, spatial autocorrelation — and that toolkit will mostly **fail**.
Understanding *why* it fails is the point of this chapter, and the reason
Chapter \@ref(rlq) needs a different method.

```{r ch2_setup, echo = FALSE, warning = FALSE, message = FALSE}
# rebuild chapter 1's outputs so this chapter can be run on its own
genome_coverage  <- breadth_of_coverage(genome_covered_bases, genome_metadata)
read_counts_filt <- filter_by_coverage(genome_read_counts, genome_coverage, 0.3)
genome_counts    <- to_genome_counts(read_counts_filt, genome_metadata)
sample_metadata  <- sample_metadata %>%
  select(-any_of(c("richness", "neutral", "phylogenetic"))) %>%
  left_join(alpha_diversity(genome_counts, genome_tree), by = "microsample")
set.seed(2024)
```

## Putting the microsamples back on the tissue

The most direct thing to do with spatial data is look at it. Every microsample
has a pixel position on the photograph of its cryosection, so any
per-microsample number can be painted onto the tissue.

```{r map_richness, warning = FALSE, message = FALSE, fig.height = 5, fig.width = 10}
plot_on_section("G121eI117A", sample_metadata, "richness",
                "Richness", pixel_coords, diverging = FALSE) +
plot_on_section("G121eO306A", sample_metadata, "richness",
                "Richness", pixel_coords, diverging = FALSE)
```

And the variable we ultimately care about — how far each microsample sits from
the epithelium, the living gut wall:

```{r map_host_distance, warning = FALSE, message = FALSE, fig.height = 5, fig.width = 10}
plot_on_section("G121eI117A", sample_metadata, "distance_host",
                "Distance to\ngut wall (µm)", pixel_coords, diverging = FALSE) +
plot_on_section("G121eO306A", sample_metadata, "distance_host",
                "Distance to\ngut wall (µm)", pixel_coords, diverging = FALSE)
```

Distance to the wall varies smoothly across the section — it is a genuine
spatial gradient, by construction. Richness looks patchier. Whether those
patches mean anything is what the rest of the chapter tests.

## Compositional data and the CLR transformation

We now have to separate two questions that the raw table mixes together: *how
much* was in a microsample, and *what it was made of*.

The first is sequencing depth, which under this protocol carries biomass
information (Chapter \@ref(data)). We set it aside deliberately here and pick it
up again as its own variable.

The second is composition, and it is genuinely compositional: within a
microsample, only the ratios between genomes are interpretable. Euclidean
distances on raw proportions are misleading, and the standard fix is the
**centred log-ratio (CLR)** — divide every value by the geometric mean of its
microsample, then take logs. Note what CLR does to the total: it divides it out
entirely. That is the right move for asking *what lives here*, and it is exactly
why the biomass signal has to be reintroduced separately later.

Logs need positive numbers, so zeros are replaced first, using a Bayesian
estimator that respects the compositional structure.

```{r clr, warning = FALSE, message = FALSE}
comm_all <- genome_counts %>%
  column_to_rownames("genome") %>%
  t() %>%
  as.data.frame()
comm_all <- comm_all[rowSums(comm_all) > 0, colSums(comm_all) > 0, drop = FALSE]

# zeros -> small positive values; genomes too sparse to estimate are dropped
comm_repl <- cmultRepl(comm_all, method = "GBM", output = "prop",
                       z.warning = 0.95)

clr <- function(x) log(x) - rowMeans(log(x))
comm_clr <- clr(as.matrix(comm_repl))

dim(comm_clr)  # microsamples x genomes retained
```

Note how much shrank: from 223 genomes to under a hundred. Everything else was
too sparse across these two cryosections to place on a log scale at all.

```{r pca, warning = FALSE, message = FALSE, fig.height = 5, fig.width = 7}
pca <- prcomp(comm_clr, center = TRUE, scale. = FALSE)
var_pct <- round(100 * pca$sdev^2 / sum(pca$sdev^2), 1)

pca_scores <- as.data.frame(pca$x[, 1:2]) %>%
  rownames_to_column("microsample") %>%
  left_join(sample_metadata, by = "microsample")

ggplot(pca_scores, aes(PC1, PC2, colour = region)) +
  geom_point(size = 2, alpha = 0.85) +
  stat_ellipse(linewidth = 0.5) +
  scale_colour_manual(values = region_colors) +
  labs(x = paste0("PC1 (", var_pct[1], "%)"),
       y = paste0("PC2 (", var_pct[2], "%)"),
       colour = "Region")
```

The two organs overlap heavily. Let us quantify that with PERMANOVA, which asks
whether a variable explains more between-sample distance than label shuffling
would produce by chance.

```{r permanova_region, warning = FALSE, message = FALSE}
dist_clr <- dist(comm_clr, method = "euclidean")   # Aitchison distance

meta_ord <- sample_metadata %>%
  filter(microsample %in% rownames(comm_clr)) %>%
  arrange(match(microsample, rownames(comm_clr)))

adonis2(dist_clr ~ region, data = meta_ord, permutations = 999)
```

Caecum and colon differ significantly — but **R² is only about 3%**. Two
different organs, and the difference between them accounts for a thirtieth of
the variation in the data.

::: {.rmdnote}
**First real result.** At this scale, variation *between neighbouring
microsamples within one section* dwarfs the variation *between gut regions*.
That is the whole justification for micro-scale sampling: homogenise a caecum
and you average away 97% of what is actually going on.
:::

## Working within a single cryosection

If local variation is where the signal is, the analysis belongs inside one
cryosection, where gut region, animal, and laboratory batch are all held
constant and the only thing that varies is position.

`prepare_section()` does that: it pulls out one section, drops empty
microsamples, and replaces zeros within that section alone.

```{r sections, warning = FALSE, message = FALSE}
caecum <- prepare_section("G121eI117A", genome_counts, sample_metadata)
colon  <- prepare_section("G121eO306A", genome_counts, sample_metadata)

tibble(
  section     = c(caecum$section, colon$section),
  microsamples = c(nrow(caecum$comm_red), nrow(colon$comm_red)),
  genomes      = c(ncol(caecum$comm_red), ncol(colon$comm_red))
)
```

Now the key test. Does distance to the gut wall explain community composition
inside a single 2 mm section?

```{r permanova_within, warning = FALSE, message = FALSE}
caecum_clr  <- clr(as.matrix(caecum$comm_zero_repl))
caecum_dist <- dist(caecum_clr)
caecum_meta <- caecum$metadata

adonis2(caecum_dist ~ distance_host, data = caecum_meta, permutations = 999)
```

**Not significant.** Now compare that with the coordinates themselves, and with
sequencing depth — in both sections:

```{r permanova_compare, warning = FALSE, message = FALSE}
colon_clr  <- clr(as.matrix(colon$comm_zero_repl))
colon_dist <- dist(colon_clr)
colon_meta <- colon$metadata

permanova_row <- function(dist_matrix, meta, rhs, label, section) {
  a <- adonis2(as.formula(paste("dist_matrix ~", rhs)),
               data = meta, permutations = 999)
  tibble(section = section, term = label,
         R2 = round(a$R2[1], 4), F = round(a$F[1], 2), p = a$`Pr(>F)`[1])
}

terms <- list(
  c("distance_host",                    "distance to gut wall"),
  c("Xcoord + Ycoord",                  "X + Y coordinates"),
  c("log(after_filtering.total_bases)", "log(sequencing depth)")
)

bind_rows(
  map_dfr(terms, ~ permanova_row(caecum_dist, caecum_meta, .x[1], .x[2], "caecum")),
  map_dfr(terms, ~ permanova_row(colon_dist,  colon_meta,  .x[1], .x[2], "colon"))
)
```

::: {.rmdnote}
**Permutation p-values wobble.** PERMANOVA estimates significance by shuffling,
so re-running the same test gives slightly different p-values (0.08 rather than
0.09, say). Read them as approximate, and never as a hard threshold.
:::

Sequencing depth is by far the strongest predictor of community composition, in
both sections — five to eleven times more explanatory than distance to the gut
wall.

In an equimolar-pooled study this would be a red flag: a technical variable
driving composition means something went wrong. Here it is not. Depth stands in
for biomass, and it is entirely reasonable that microsamples holding more
bacteria hold a *different* community from microsamples holding very few. Dense
and sparse patches of gut content are different habitats.

What the result does mean is that biomass, not position, is the leading axis of
compositional variation inside a cryosection. Any spatial gradient we hope to
find has to be looked for underneath it.

## Spatial autocorrelation

To measure spatial structure directly we need a definition of "nearby".
Microsamples are not on a regular grid — the laser cuts where the tissue allows
— so rather than pick a distance cutoff we build a **Gabriel graph**: two
microsamples are joined when no third one lies inside the circle having them as
its diameter.

```{r neighbours, warning = FALSE, message = FALSE, fig.height = 5, fig.width = 6}
caecum_nb <- neighbour_graph(caecum_meta[, c("Xcoord", "Ycoord")])
caecum_nb

plot_neighbour_graph(caecum_meta[, c("Xcoord", "Ycoord")], caecum_nb,
                     point_values = caecum_meta$richness,
                     value_label = "Richness") +
  labs(title = "Gabriel neighbour graph, caecum section G121eI117A")
```

**Moran's I** is a correlation coefficient for space: it compares each
microsample's value with its neighbours'.

- **I > 0** — neighbours resemble each other (spatial structure)
- **I ≈ 0** — values arranged at random
- **I < 0** — neighbours systematically *unlike* each other

Computing it at increasing lag orders — direct neighbours, then
neighbours-of-neighbours, and so on — shows how far the resemblance reaches.

```{r correlogram, warning = FALSE, message = FALSE, fig.height = 4, fig.width = 9}
correlograms <- bind_rows(
  moran_correlogram(caecum_nb, log(caecum_meta$distance_host)) %>%
    mutate(variable = "log(distance to gut wall)"),
  moran_correlogram(caecum_nb, log(caecum_meta$after_filtering.total_bases)) %>%
    mutate(variable = "log(sequencing depth)"),
  moran_correlogram(caecum_nb, log(caecum_meta$richness)) %>%
    mutate(variable = "log(richness)"),
  moran_correlogram(caecum_nb, prcomp(caecum_clr)$x[, 1]) %>%
    mutate(variable = "composition (CLR-PCA axis 1)")
) %>%
  mutate(variable = fct_inorder(variable))

ggplot(correlograms, aes(x = lag, y = morans_i)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "#4a6ab7", alpha = 0.2) +
  geom_line(colour = "#4a6ab7") +
  geom_point(colour = "#4a6ab7", size = 1.8) +
  facet_wrap(~ variable, nrow = 1) +
  scale_x_continuous(breaks = seq(2, 8, 2)) +
  labs(x = "Lag order (steps through the neighbour graph)", y = "Moran's I")
```

Read the four panels in order:

1. **Distance to the gut wall** is strongly autocorrelated at short lags and
   turns negative at long ones — the signature of a gradient. This is
   guaranteed: it is a smooth function of position.
2. **Sequencing depth** is also autocorrelated, weakly but clearly (I = 0.14,
   p = 0.003). Read as biomass, this is a result in its own right: **microbial
   biomass is spatially patchy at the sub-millimetre scale.** Neighbouring
   pieces of tissue tend to be densely or sparsely colonised together.
3. **Richness** is essentially flat — despite correlating with depth at ρ > 0.9.
   Worth pausing on: two variables can be tightly correlated and still differ in
   spatial behaviour.
4. **Composition**, summarised as the first CLR-PCA axis, is **also flat**.

```{r moran_tests, warning = FALSE, message = FALSE}
lw <- nb2listw(caecum_nb)

map_dfr(
  list(
    "log(distance to gut wall)" = log(caecum_meta$distance_host),
    "log(sequencing depth)"     = log(caecum_meta$after_filtering.total_bases),
    "log(richness)"             = log(caecum_meta$richness),
    "composition (PCA axis 1)"  = prcomp(caecum_clr)$x[, 1]
  ),
  function(v) {
    tst <- moran.test(v, lw, zero.policy = TRUE)
    tibble(morans_i = round(unname(tst$estimate[1]), 3),
           p_value  = signif(tst$p.value, 3))
  },
  .id = "variable"
)
```

## Why the obvious approach failed

By the standard toolkit, this dataset has **no spatial structure in community
composition**. But that conclusion rests entirely on one choice we never
examined: using PCA axis 1 as the summary of "what lives here".

PCA ranks its axes by **variance**, not by relevance. Nothing guarantees that
the largest axis is the interesting one. Look at what the first three axes
actually correspond to:

```{r pca_axes_meaning, warning = FALSE, message = FALSE}
caecum_pca <- prcomp(caecum_clr)
caecum_var <- round(100 * caecum_pca$sdev^2 / sum(caecum_pca$sdev^2), 1)

map_dfr(1:3, function(k) {
  ax  <- caecum_pca$x[, k]
  tst <- moran.test(ax, lw, zero.policy = TRUE)
  tibble(
    axis            = paste0("PC", k),
    variance_pct    = caecum_var[k],
    rho_depth       = round(cor(ax, log(caecum_meta$after_filtering.total_bases),
                                method = "spearman"), 2),
    rho_host_dist   = round(cor(ax, log(caecum_meta$distance_host),
                                method = "spearman"), 2),
    morans_i        = round(unname(tst$estimate[1]), 3),
    moran_p         = signif(tst$p.value, 3)
  )
})
```

Read that table carefully — it is the turning point of the practical.

- **PC1** is the biggest axis (13% of variance) and corresponds to **nothing we
  measured**: not depth, not position, and it is not spatially autocorrelated.
  It is unstructured microsample-to-microsample turnover — the stochastic part
  of who happens to be where.
- **PC2** is the biomass axis (ρ ≈ 0.8 with depth) and *is* spatially
  autocorrelated, consistent with the patchiness we saw above.
- **PC3** carries the weak trace of distance to the gut wall (ρ ≈ 0.2) and is
  also significantly autocorrelated.

**The gradient we came for is on axis 3.** It was never going to be found by
looking at axis 1, and it is smeared thinly across several axes rather than
isolated on any of them — because nothing in PCA's objective function has any
reason to isolate it.

::: {.rmdnote}
**What we need instead** is a method that does not rank axes by size, but looks
for the axis that is simultaneously **organised in space** and **associated with
the kind of bacteria involved** — their metabolism and their evolutionary
history. Sequencing noise cannot fake that combination: there is no reason for a
microsample's biomass to correlate with what its bacteria can metabolise. That
is exactly what extended RLQ does, and it is the subject of
Chapter \@ref(rlq).
:::
