# Exercises {#exercises}

Every exercise below can be answered with the objects already built in
Chapters \@ref(data)–\@ref(rlq). Run the three chapters first, or paste their
setup chunks into a fresh session.

Exercises marked **★** are the ones worth doing if time is short.

## Set 1 — The coverage filter {-}

The 0.3 breadth-of-coverage threshold was presented as a given. It is a choice.

**1.1 ★** Rebuild the community table at `min_coverage` values of 0, 0.1, 0.3
and 0.6. For each, record how many non-zero entries survive, what fraction of
reads is kept, and the median richness per microsample.

```{r ex_1_1, eval = FALSE}
map_dfr(c(0, 0.1, 0.3, 0.6), function(threshold) {
  filtered <- filter_by_coverage(genome_read_counts, genome_coverage, threshold)
  counts   <- to_genome_counts(filtered, genome_metadata)
  tibble(
    threshold       = threshold,
    nonzero_entries = sum(filtered[, -1] > 0),
    reads_kept      = sum(filtered[, -1]) / sum(genome_read_counts[, -1]),
    median_richness = median(colSums(counts[, -1] > 0))
  )
})
```

**1.2** At threshold 0, richness is roughly 220 genomes per microsample — nearly
the whole catalogue. Why is that not a plausible biological result for a piece
of tissue a few hundred micrometres across?

**1.3** Re-run the caecum RLQ (Chapter \@ref(rlq)) using a table filtered at 0.1
and again at 0.6. Does axis 2 still correlate with distance to the gut wall? Is
the conclusion robust to the threshold, or does it depend on it?

## Set 2 — Compositionality {-}

**2.1** Chapter \@ref(space) replaced zeros before taking logs. Try instead the
common shortcut of adding a pseudocount of 1 to every value, then CLR-transform
and redo the PCA. How much does the ordination change? Which genomes move most,
and what do they have in common?

**2.2 ★** `cmultRepl()` dropped genomes present in fewer than 5% of
microsamples. Those are rare taxa. Argue both sides: what is lost by discarding
them, and what would go wrong if we kept them?

**2.3 ★** CLR divides out each microsample's total, and Chapter \@ref(space)
argued that this is why biomass has to be handed to RLQ separately. Test the
claim directly: compute each microsample's total genome count, and check whether
it survives the CLR transformation in any recoverable form.

```{r ex_2_3, eval = FALSE}
totals <- rowSums(caecum$comm_red)

# does anything in the CLR matrix still track the total?
cor(rowSums(caecum_clr), log(totals))          # CLR rows sum to zero by construction
max(abs(cor(caecum_clr, log(totals))))         # any single genome?
```

Given the answer, would an analysis that used *only* CLR composition be able to
detect a biomass gradient at all?

## Set 3 — Spatial structure {-}

**3.1 ★** The Gabriel graph was one choice of neighbourhood. Rebuild it as a
k-nearest-neighbour graph and as a distance-band graph:

```{r ex_3_1, eval = FALSE}
coords <- as.matrix(caecum_meta[, c("Xcoord", "Ycoord")])

nb_gabriel <- neighbour_graph(coords)
nb_knn     <- knn2nb(knearneigh(coords, k = 4))
nb_dist    <- dnearneigh(coords, 0, 400)   # within 400 µm

# how many links does each definition produce?
map_dbl(list(nb_gabriel, nb_knn, nb_dist), ~ sum(card(.x)))
```

Recompute Moran's I for distance-to-gut-wall under all three. Does the
conclusion depend on the neighbourhood definition? Should it?

**3.2 ★** In Chapter \@ref(space), sequencing depth was spatially
autocorrelated in the caecum (I = 0.14) but not in the colon (I = 0.05, n.s.).
Under the unnormalised pooling protocol this reads as *biomass is spatially
patchy in the caecum*. Give one competing explanation that is purely technical,
and say what you would measure to tell them apart.

**3.3** The microsamples of one cryosection were prepared in four batches
(`batch`, 42 microsamples each). If batches were laid out spatially rather than
at random, batch effects would masquerade as biomass patchiness. Check whether
`batch` is spatially clustered on the section, and whether sequencing depth
differs between batches.

**3.4 ★** Repeat the shuffled-coordinates null model, but instead of shuffling
freely, shuffle *within* bands of distance to the gut wall. What does the
comparison between the two nulls tell you that either alone does not?

## Set 4 — RLQ {-}

**4.1 ★** Table **E** contained three variables: sequencing depth, richness and
distance to the gut wall. Drop distance to the gut wall from `spatial_rlq()` and
re-run the caecum analysis. Does axis 2 still recover the spatial gradient?

This is the most important exercise in the practical. If the gradient survives,
RLQ found it through **S** (the neighbourhood structure) and **L** alone, which
means it is not an artefact of having told the method what to look for.

```{r ex_4_1, eval = FALSE}
# edit the `env` data frame inside spatial_rlq(), or copy the function and
# remove the log_host_dist column, then compare:
#   cor(new_sites$axis2, log(new_sites$distance_host), method = "spearman")
```

**4.2** RLQ was given the phylogeny (**P**) *and* metabolic traits (**T**), and
Chapter \@ref(rlq) showed they carry overlapping information. Re-run with only
one of them. Which matters more for recovering the gradient?

**4.3 ★** Caecum axis 1 tracked sequencing depth almost perfectly (ρ ≈ 0.93),
and the practical read that as a **biomass** axis rather than a technical one —
because libraries were pooled without normalisation.

That interpretation is an argument about protocol, not about the data in front
of you. Set out what the data could show that would *contradict* it. Then say
whether any of those checks is possible with the variables in
`sample_metadata`, and what extra measurement would settle the question
outright.

**4.4 ★** The colon section showed no gradient. List three explanations that
have nothing to do with the colon lacking spatial structure, and say how you
would distinguish them.

## Set 5 — Interpretation {-}

**5.1 ★** The concluding table of Chapter \@ref(rlq) claims that genomes further
from the gut wall are larger and better equipped for polysaccharide degradation.
Genome size and polysaccharide-degradation capacity are themselves correlated —
bigger genomes encode more of everything. Design an analysis that separates the
two explanations.

**5.2** Blomberg's *K* was 1.5 (p = 0.001): spatial preference is
phylogenetically conserved. Does that support the metabolic explanation,
undermine it, or neither?

**5.3** You have one animal and one cryosection per gut region. Write the
strongest version of the criticism a reviewer would make, and state what the
study would need to answer it.

## Set 6 — Extension {-}

**6.1** The published study analysed six cryosections — three caecum, three
colon — and asked whether the same gradient reappeared in each. Describe how you
would test "the same gradient" quantitatively, given that RLQ axes from separate
analyses are not directly comparable (different genomes, arbitrary signs,
different scaling).

**6.2** Microsamples here are 5,000 µm². If you could re-run the experiment at a
quarter of that area, with four times as many microsamples and the same total
sequencing cost, would you? Argue it in terms of what Chapter \@ref(space)
showed about sequencing depth — and remember that under unnormalised pooling the
faintest microsamples receive very few reads rather than a guaranteed share.

**6.3 ★** Suppose a reviewer insists you pool equimolarly, as standard practice
requires. Write the paragraph you would send back. What is lost, what is gained,
and is there a design that gets both?

---

## Where to go next {-}

The full analysis behind the manuscript — method validation against FISH
microscopy and conventional metagenomics, genome-catalogue comparison,
contamination assessment, strain-level microdiversity, and the six-cryosection
replication — is at
[github.com/3d-omics/MSSM](https://github.com/3d-omics/MSSM).

**Key references**

- Pavoine S., Rodríguez-Gironés M.A. & Cabana J. (2011) Extended RLQ: linking
  environment, space, traits and phylogeny. *Journal of Ecology*.
- Dray S. & Legendre P. (2008) Testing the species traits–environment
  relationships: the fourth-corner problem revisited. *Ecology*.
- Aitchison J. (1986) *The Statistical Analysis of Compositional Data*.
- Gloor G.B. *et al.* (2017) Microbiome datasets are compositional: and this is
  not optional. *Frontiers in Microbiology*.
