---
title: "Micro-scale spatial metagenomics | Practical"
author: "AlberdiLab · 3D'omics"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
url: "https://github.com/3d-omics/MSSM-practical"
description: |
  A teaching version of the micro-scale spatial metagenomics analysis:
  the minimum set of steps needed to understand how spatial structure in a
  gut microbiome is measured and tested.
link-citations: true
---

```{r knitr_opts, echo = FALSE}
knitr::opts_chunk$set(
  class.source = "script-source",
  class.output = "script-output",
  comment = NA,
  fig.align = "center"
)
```

# Introduction

Conventional metagenomics grinds up a whole gut sample and asks *which bacteria
live in this animal*. It cannot ask **where** they live, because the moment the
sample is homogenised all spatial information is gone.

**Micro-scale spatial metagenomics (MSSM)** keeps that information. A frozen
section of intestinal tissue is photographed, and a laser cuts out hundreds of
tiny pieces — *microsamples* — each only a few hundred micrometres across. Each
microsample is sequenced separately, and because the laser recorded where it
cut, every community profile comes with an **X/Y coordinate on the tissue**.

The data therefore has a shape that ordinary microbiome data does not:

```
  microsample  ──  community composition   (which genomes, how much of each)
       │
       ├─────────  position in the tissue  (X, Y, distance to the gut wall)
       └─────────  sequencing metrics      (depth, diversity)

  genome       ──  taxonomy, genome size, phylogeny, metabolic capacity
```

This practical works through the analysis that data structure makes possible,
using real data from one chicken caecum and one chicken colon.

## Study design

```{r design_figure, echo = FALSE, out.width = "100%", fig.cap = "How MSSM data is produced, from tissue to the five tables that enter the analysis."}
knitr::include_graphics("resources/images/study_design.svg")
```

Steps 1, 2 and 4 are what you would expect. **Step 3 is not**, and the next
section explains why it matters more than it looks.

## The question

> Bacteria in the gut are not randomly mixed. Communities a few hundred
> micrometres apart differ, and they differ *systematically* — along gradients
> that run from the gut wall out into the lumen. Can we detect such a gradient,
> and can we say which kind of bacteria sit at each end of it?

Answering it takes four steps, one per chapter:

| Chapter | Question | Method |
|---|---|---|
| \@ref(data) | What does a microsample community actually look like? | Coverage filtering, genome counts, Hill numbers |
| \@ref(space) | Is there *any* spatial structure, or is it noise? | Tissue maps, CLR + PCA, PERMANOVA, Moran's I |
| \@ref(rlq) | What explains the structure? | Extended RLQ (space + environment + traits + phylogeny) |
| \@ref(exercises) | Does it hold up? | Exercises: null models, replication, parameter choices |

## The data

Everything you need is in this repository — nothing to download.

| Object | What it is |
|---|---|
| `sample_metadata` | 336 microsamples: coordinates, distance to gut wall, sequencing depth |
| `genome_read_counts` | Reads mapped to each of 223 genomes in each microsample |
| `genome_covered_bases` | How much of each genome those reads actually covered |
| `genome_metadata` | Taxonomy, genome length, completeness of each genome |
| `genome_tree` | Phylogeny of the 223 genomes |
| `genome_gifts` | Metabolic capacity of each genome (GIFT scores) |

The 223 genomes are metagenome-assembled genomes (MAGs) built beforehand from
conventional, macro-scale sequencing of the same animals. Microsample reads are
mapped against that catalogue — a microsample on its own has far too little DNA
to assemble anything.

Two cryosections are included, chosen so you can compare gut regions:

| Cryosection | Region | Microsamples |
|---|---|---|
| `G121eI117A` | Caecum | 168 |
| `G121eO306A` | Colon | 168 |

Both come from the same animal (G121), so differences between them are about
gut region and not about which chicken they came from.

### A deliberately unnormalised protocol

One design decision shapes every analysis in this book, and it is worth
understanding before you touch the data.

In conventional metagenomics, each library is quantified after preparation and
pooled **equimolarly**, so that every sample receives roughly the same number of
reads. Sequencing depth is then a technical nuisance, deliberately equalised,
and any residual variation in it is something to correct for.

MSSM does the opposite:

- every microsample is cut to the **same area**, so the same amount of tissue
  enters each library;
- every library gets the **same number of PCR cycles**, so amplification does
  not rescale anything;
- libraries are pooled **without normalisation**;
- all libraries from one cryosection are sequenced in a **single run**.

The consequence is that the reads a microsample receives are roughly
proportional to the DNA that went into it — and therefore to the **microbial
biomass it contained**. Depth stops being a nuisance and becomes a
quasi-quantitative measurement that equimolar pooling would have destroyed.

Two caveats come with it, and both matter later:

1. **It is relative, not absolute.** One run is shared out among the
   microsamples of one cryosection, so a microsample's depth is its *share* of
   that cryosection's total biomass. The composition closes at the cryosection
   level, and depths are not comparable between cryosections or runs. This is
   why every analysis in Chapters \@ref(space) and \@ref(rlq) is done within a
   single section.
2. **It is a proxy, not a calibrated measurement.** Extraction efficiency,
   library quality and PCR stochasticity vary too. And because deeper
   sequencing detects more genomes, anything computed from depth — richness
   above all — inherits both the biological signal and the detection effect.

::: {.rmdnote}
**Read this as an unusual choice, not an oversight.** For most metagenomics,
unnormalised pooling would be a mistake: it wastes reads on samples that do not
need them and starves those that do. For MSSM it is the point. Biomass varies
enormously between neighbouring pieces of tissue, that variation is part of the
spatial biology being studied, and equimolar pooling would erase it.
:::

::: {.rmdnote}
**This is a simplified version.** The published study also validated the method
against FISH microscopy and conventional metagenomics, compared genome
catalogues, quantified contamination, and analysed strain-level variation. None
of that is here. The full analysis lives at
[github.com/3d-omics/MSSM](https://github.com/3d-omics/MSSM).
:::

## Setting up

### Packages

```{r load_libraries, warning = FALSE, message = FALSE}
if (!require("pacman")) install.packages("pacman")

# CRAN
pacman::p_load(
  tidyverse,      # data manipulation and plotting
  ape,            # phylogenetic trees
  phytools,       # tree utilities
  vegan,          # ecology: ordination, PERMANOVA, transformations
  zCompositions,  # zero replacement for compositional data
  ade4,           # the ordinations that RLQ is built from
  adespatial,     # spatial ecology
  adegraphics,    # required by ade4 for plotting
  spdep,          # neighbour graphs and Moran's I
  patchwork,      # combining plots
  ggrepel,        # non-overlapping labels
  magick          # reading the cryosection photographs
)

# Bioconductor: phylogeny plotting
# BiocManager::install("ggtree")
pacman::p_load(ggtree)

# GitHub: Hill-number diversity and metabolic capacity annotation
pacman::p_load_gh("anttonalberdi/hilldiv2", "anttonalberdi/distillR")

select <- dplyr::select
```

### Colours and theme

One palette for bacterial orders, used consistently everywhere, so a colour
means the same thing in every figure of this book.

```{r plotting_settings, warning = FALSE, message = FALSE}
phylum_colors <- c(
  Actinomycetota    = "#346254",
  Bacillota         = "#4a6ab7",
  Bacillota_A       = "#8c1c47",
  Bacteroidota      = "#9c8464",
  Pseudomonadota    = "#c49d4b",
  Verrucomicrobiota = "#462410"
)

order_colors <- c(
  Actinomycetales      = "#79b1a3",
  Coriobacteriales     = "#498a77",
  CAJFEE01             = "#a0c5e8",
  Erysipelotrichales   = "#7acef4",
  Lactobacillales      = "#6390fb",
  ML615J_28            = "#70c9de",
  Paenibacillales      = "#70a8dc",
  RF39                 = "#5c7fba",
  Christensenellales   = "#f58262",
  Clostridiales        = "#cf4a82",
  Lachnospirales       = "#e15f7d",
  Monoglobales         = "#e18299",
  Oscillospirales      = "#ba2760",
  Peptostreptococcales = "#e35d51",
  TANB77               = "#ef3d26",
  UBA1381              = "#be5643",
  Bacteroidales        = "#e3c7a0",
  Enterobacterales     = "#ffcc62",
  Verrucomicrobiales   = "#823f1d",
  Mycobacteriales      = "#79b1b6",
  Propionibacteriales  = "darkseagreen3",
  Azospirillales       = "lightyellow",
  Eubacteriales        = "maroon"
)

region_colors <- c(Caecum = "#2A9D8F", Colon = "#E76F51")

mssm_theme <- theme_classic(base_size = 12) +
  theme(
    plot.title   = element_text(size = 12, face = "bold"),
    legend.title = element_text(size = 10),
    legend.text  = element_text(size = 9),
    strip.background = element_rect(fill = "#dde3e9", colour = "white")
  )

theme_set(mssm_theme)
```

### Helper functions

The steps that would otherwise be twenty lines of bookkeeping are wrapped in
functions. Open
[`resources/scripts/practical_functions.R`](resources/scripts/practical_functions.R)
whenever you want to see what one of them really does — each has a comment
explaining the reasoning, not just the mechanics.

```{r load_functions, warning = FALSE, message = FALSE}
source("resources/scripts/rlqESLTP.R")           # extended RLQ (Pavoine et al. 2011)
source("resources/scripts/practical_functions.R") # everything else
```

### Loading the data

```{r load_data, warning = FALSE, message = FALSE}
load("resources/data/mssm_practical_data.Rdata")

pixel_coords <- read_csv("resources/images/microsample_pixel_coords.csv",
                         show_col_types = FALSE)

ls()
```
